Bootstrapping Multi-atlas Hippocampal Segmentation with MAGeT

### 2013-02-15 paper.Rnw
Updated methods

Multi-atlas segmentation techniques address these limitations by combining
segmentation information from a series of expertly segmented atlases
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2012,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used merge
these labels into a definitive segmentation for the target.

Multi-atlas methods have been very successfully applied to hippocampal
segmentation. Collins et al. found near-manual segmentation performance using an
atlas library of 80 T1-weighted atlas images from the ICBM152 dataset, the
ANIMAL nonlinear registration algorithm, normalised mutual information as a
similarity metric for atlas selection, and majority vote for label fusion
\citep{Collins2010}.

The Alzheimer's Disease Neuroimaging Initiative (ADNI) is a
commonly used benchmarking dataset of MR images of controls and patients with
MCI or Alzheimer's (see Methods for more information on the ADNI dataset).

Leung et al. tuned parameters for registration and label fusion to the
segmentation of ADNI1 1 year dataset of images with an atlas library of 55
images \citep{Leung2010}.

The MAPER whole brain segmentation algorithm \citep{Heckemann2006,Heckemann2011},
using 30 atlases, on all ADNI1 baseline images \citep{Heckemann2011}.

In \citep{Lotjonen2012}, the authors use a proprietary non-linear registration
method based on intensity differences, and post-processing step using a graph
cuts or EM approach with intensity spatial prior models to optimise the fused
segmentations.  Using images from the ADNI1 baseline dataset and 30 atlases.

{\em Atlas selection} techniques are often used to filter out atlases dissimilar
to a target image in order to reduce {conceptual} errors \citep{Aljabar2009}.

The LEAP algorithm, Wolz et al.  is an elegant modification to the basic
multi-atlas strategy: the atlas library is grown, beginning with a set of
manually labelled atlases and successively incorporating unlabelled target
images after being labelling using multi-atlas techniques\citep{Wolz2010}.  The
sequence in which target images are labelled is chosen so that the similarity
between the atlas images and the target images is minimised at each step,
effectively allowing for deformations between very dissimilar images to be
broken up into sequences of smaller deformations.  With an atlas library of 30
MR images, LEAP was used to segment the ADNI1 baseline dataset, achieving a mean
Dice score of 0.85 with manual segmentations.

@@

While not purely nulti-atlas techniques, there are several important algorithms
for hippocampal segmentation that inform our approach. The popular FreeSurfer
application's whole brain segmentation algoritm uses a probabilistic atlas of
anatomical and tissue classes along with spatial constraints for class labels
encoded using a Markov random field model \citep{Fischl2002}. When segmenting
hippocampal subfields, FreeSurfer employs a Bayesian inference algorithm using a
probabilistic atlas of anatomical classes as a prior, and a likelihood model of
how those classes translate into MR image intensities, both trained on manual
segmentations of high resolution MR images \citep{Fischl2009}. Yushkevitch et
al. describe a semi-automated method for hippocampal subfield segmentation of
focal T2 images\citep{Yushkevitch2010}.  The unlabelled MR image must be
manually partitioned into 'head', 'body' and 'tail', and then multi-atlas
methods are used to segment the image.  Finally, an AdaBoost-based bias
correction classifier is trained on texture, spatial location, and intensities
of manual segmentations and is applied to fix mislabelled voxels.  
%  Chupin's use of shape-based priors 
%   - Mouiha2011b - Hippocampal atrophy rates in AD: compare SNT vs FreeSurfer
%     on ADNI
%

Aside from the algorithmic choices used in multi-atlas segmentation, it is
natural to ask about how the features of the atlases themselves impact the
resulting segmentations. As noted, by choosing atlases ranked most similar to a
target image by voxel intensity profile segmentation accuracy is improved,
suggesting that neuroanatomical similarity plays a role\citep{Aljabar2009}.
Carmichael et al. explored this directly and found that when using only one
atlas the important factors leading to improved accuracy are that the atlas have
neuroanatomical features that match the target, and that the atlas segmentation
use the same protocol as the gold-standard \citep{Carmichael2005}. Nestor et al.
found that hippocampal segmentation protocols that include more dorsal
white-matter and posterior anatomy tended to produce higher overlap and better
accuracy at distinguishing disease classes in the ADNI1 1 year dataset
\citep{Nestor2012}. These results suggest both atlas library neuroanatomy and
delineation protocol play a significant role in the resulting segmentation.

@@

\subsubsection{SZ First Episode Patients}
Nomenclature: SZFEP Dataset
SZFEP dataset demographics are shown in table \ref{tab:SZFEP_Demographics}.

\todo{include image characteristics}

@@

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI1 cross-validation}
To test the effect of parameter settings on the MAGeT brain algorithm, repeated
random sub-sampling cross-validation (RRSCV) was performed using a subset of the
ADNI1 subjects for validation.  This form of cross-validation (described in
detail below) allows us to validate MAGeT with various atlas and template
library sizes, registration and label fusion methods.

@@

{\bf Evaluation.}  Repeated random sub-sampling cross-validation (RRSCV)
consists of repeated trials in which items from the dataset are randomly sampled
and assigned to a training set or validation set. In each trial, performance on
the validation set is measured, and then averaged across all trials. This form
of cross-validation allows

With respect to this We performed RRSCV on each combination of parameters listed above: atlas library
size, template library size, registration method, and label fusion method.

We performed 10 trials per parameter combination.  In each validation trial, the
training set consisted of the images used as atlases, and the validation set
consisted of the images to be segmented.

The MAGeT brain algorithm and the basic multi-atlas segmentation procedure were
applied to segment the images in the validation set.  Additionally, in each
trial, the single-atlas segmentation was obtained for each atlas-template.

The gold-standard for the segmentation accuracy of images in the validation set
was the SNT manual labels.

\todo{ show off the number of registrations/comparisions we did}

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
\todo{comparison across field strengths, a la Heckemann2011?}
To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

@@

{\bf Atlas and template libary.} The atlas library consisted of the entire 
Winterburn atlas set.  The Winterburn atlases are digital segmentations of the
hippocampus in five in-vivo 300u isotropic T1-weighted MR scans, and include 
subfield segmentations for the cornus ammonis (CA) 1, CA4, dentate gyrus, 
subiculum, and CA 2 and 3 combined. Subjects in the Winterburn atlases range in 
age from 29-57 years (mean age of 37), and include two males and three females.

The template library consisted of 21 randomly selected images from the ADNI1 
data dataset (7 healthy, MCI and AD subjects).

{\bf Registration method.} ANTS, as it performed best in the cross-validation
experiment.

\todo{okay to reference results whilst still in Methods?}

{\bf Label fusion.} Majority vote, as it is simpliest to run and performed 
equally well in cross-validation experiment.

{\bf Evaluation.}

Since hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct similarity comparisons
between labels produced by MAGeT brain and the ADNI labels.

@@

To evaluate the performance of MAGeT brain, we correlate our segmentation
volumes with manual segmentation volumes, as well as with hippocampal volumes
of established automated segmentation methods.

Additionally, we compared classification accuracy of subjects
by diagnosis based on hippocampal volume using both the SMT labels and our
produced labels.

@@

\todo{Also consider self-consistency (i.e. Kappa of same subject at different
field strengths) a la Heckemann2011}

@@

{\bf Dataset evaluated.} To validate that MAGeT performance generalises to other
diseases, we measure the performance using the best parameter settings previous
found, on a dataset consiting of first episode schizophrenia patients.
\todo{Refer to the FEP description above... or just include it here}.

{\bf Atlas and template libary.} - two different atlas sets: a manual
hippocampal segmentation of patients, and Winterburn atlas set.

{\bf Registration method.} ANTS.

{\bf Label fusion.} Majority vote.

{\bf Evaluation.} We validate the FEP-atlas segmentations using Dice's Kappa,
and the Winterburn-atlas segmentations by correlating volumes.

%%%%%
%%%%%              Experiments: Winterburn Atlases Validation              %%%%%
%%%%%
\subsubsection{Winterburn Atlases Validation}
{\bf Dataset evaluated.} - T1 BRAVO scans of the same subjects included in the 
Winterburn atlas set.  These scans are taken within .... weeks of the scans for 
the Winterburn atlases.

{\bf Atlas and template libary.} - Atlas library is Winterburn T1 atlases.  
Template library consists of all five T1 BRAVOs, plus 15 T1 healthy control images.

{\bf Registration method.} ANTS

{\bf Label fusion.} Majority vote.

@@

Segmentation accuracy is judged by difference in hippocampal volume.

>>>>

Multi-atlas segmentation techniques address these limitations by combining
segmentation information from a series of expertly segmented atlases
\citep{Heckemann2006,Heckemann2011,Collins2010,Lotjonen2012,Aljabar2009,Leung2010,Wolz2010}.
Each atlas image is registered to a target image, and label propagation is
performed to produce several labellings of the target image (one from each
atlas). A {\it label fusion} technique, such as voxel-wise voting, is used merge
these labels into a definitive segmentation for the target.  In addition, {\em
atlas selection} techniques are often used to exclude atlases from label fusion
that are dissimilar to a target image in order to reduce error from
unrepresentative anatomy \citep{Aljabar2009}. Cross-correlation or normalised
mutual information of image intensities are common measures of image similarity
used in atlas selection.

Multi-atlas methods have been very successfully applied to hippocampal
segmentation. Collins et al. found near-manual segmentation performance using an
atlas library of 80 T1-weighted atlas images from the ICBM152 dataset, the
ANIMAL nonlinear registration algorithm, normalised mutual information as an
atlas selection similarity metric, and majority vote for label fusion
\citep{Collins2010}.  The Alzheimer's Disease Neuroimaging Initiative (ADNI) is
a commonly used benchmarking dataset of MR images of controls and patients with
MCI or Alzheimer's (see Methods for more information on the ADNI dataset).
Leung et al. tuned parameters for registration and label fusion to the
segmentation of ADNI1 1 year dataset of images with an atlas library of 55
images \citep{Leung2010}.  The MAPER whole brain segmentation algorithm
\citep{Heckemann2006,Heckemann2011}, using 30 atlases, on all ADNI1 baseline
images \citep{Heckemann2011}.  Lotjonen et al. use images from the ADNI1
baseline dataset and 30 atlases with a proprietary non-linear registration
method based on intensity differences, and post-processing step using a graph
cuts algorithm to optimise fused segmentations against a spatial intensity prior
\citep{Lotjonen2012}.  
\todo{No story here}

The LEAP algorithm is an elegant modification to the basic multi-atlas strategy
\citep{Wolz2010}. The atlas library is grown, beginning with a set of manually
labelled atlases, and successively incorporats unlabelled target images after
being labelling using multi-atlas techniques. The sequence in which target
images are labelled is chosen so that the similarity between the atlas images
and the target images is minimised at each step, effectively allowing for
deformations between very dissimilar images to be broken up into sequences of
smaller deformations.  With an atlas library of 30 MR images, LEAP was used to
segment the ADNI1 baseline dataset, achieving a mean Dice score of 0.85 with
manual segmentations.

@@

While not purely nulti-atlas techniques, there are several important algorithms
for hippocampal segmentation that inform our approach. The popular FreeSurfer
application's whole brain segmentation algoritm uses a probabilistic atlas of
anatomical and tissue classes along with spatial constraints for class labels
encoded using a Markov random field model \citep{Fischl2002}. When segmenting
hippocampal subfields, FreeSurfer employs a Bayesian inference algorithm using a
probabilistic atlas of anatomical classes as a prior, and a likelihood model of
how those classes translate into MR image intensities, both trained on manual
segmentations of high resolution MR images \citep{Fischl2009}. Yushkevitch et
al. describe a semi-automated method for hippocampal subfield segmentation of
focal T2 images\citep{Yushkevitch2010}.  The unlabelled MR image must be
manually partitioned into 'head', 'body' and 'tail', and then multi-atlas
methods are used to segment the image.  Finally, an AdaBoost-based bias
correction classifier is trained on texture, spatial location, and intensities
of manual segmentations and is applied to fix mislabelled voxels.

%  Chupin's use of shape-based priors 
%   - Mouiha2011b - Hippocampal atrophy rates in AD: compare SNT vs FreeSurfer
%     on ADNI
%

Aside from the algorithmic choices used in multi-atlas segmentation, it is
natural to ask about how the features of the atlases themselves impact the
resulting segmentations. As noted, by choosing atlases ranked most similar to a
target image by voxel intensity profile, segmentation accuracy is improved,
suggesting that neuroanatomical similarity plays a strong role
\citep{Aljabar2009}.  Carmichael et al. explored this directly and found that
when using only one atlas the important factors leading to improved accuracy are
that the atlas have neuroanatomical features that match the target, and that the
atlas segmentation use the same protocol as the gold-standard
\citep{Carmichael2005}. Nestor et al.  found that hippocampal segmentation
protocols that include more dorsal white-matter and posterior anatomy tended to
produce higher overlap and better accuracy at distinguishing disease classes in
the ADNI1 1 year dataset \citep{Nestor2012}. These results suggest both atlas
library neuroanatomy and delineation protocol play a significant role in the
resulting segmentation.

@@

\subsubsection{Schizophrenic First Episode Patients}
\todo{Nomenclature: SFEP Dataset}
\todo{describe image characteristics}
\todo{describe expert segmentations}
SFEP dataset demographics are shown in table \ref{tab:SZFEP_Demographics}.

@@

%%%%%
%%%%%                 Experiments: ADNI-1 cross-validationn               %%%%%
%%%%%
\subsubsection{ADNI1 cross-validation}
To test the effect of parameter settings on the MAGeT brain algorithm, Monte
Carlo Cross-Validation (MCCV) \citep{Shao1993} was performed using a subset of
the ADNI1 subjects for validation.  This form of cross-validation (described in
detail below) allows us to validate MAGeT with various atlas and template
library sizes, registration and label fusion methods.

@@

{\bf Evaluation.}  Monte Carlo Cross-Validation (MCCV), also known as repeated
random sub-sampling cross-validation, consists of repeated trials in which items
from the dataset are randomly sampled (without replacement) and assigned to a
training set or validation set \cite{Shao1993}.

In each trial, the training set consists of images and SNT labels used as
atlases, and the validation set consists of images to be segmented along with
the corresponding SNT labels used as gold-standard segmentations. Both MAGeT
brain and basic multi-atlas segmentation were performed in each trial, and the
accuracy of the resulting segmentations were measured against the SNT labels.

We performed 10 trials on each combination of parameters listed above: atlas
library size (1-9), template library size (1-20), registration method (ANTS vs.
ANIMAL), and label fusion method (MV, XC-WV, NMI-WV). A total of 1080 parameter
settings were evaluated over a total of 10800 trials.

%%%%%
%%%%%               Experiments: ADNI-1 Screening Validation               %%%%%
%%%%%
\subsubsection{ADNI-1 Screening Validation}
To test the accuracy of MAGeT brain on a real-world task we segment the entire
ADNI-1 dataset using an atlas set that is not representative of the target set.

@@

{\bf Atlas and template libary.} The Winterburn atlases were used as the atlas
library.  The template library consisted of 21 randomly selected images from the
ADNI1 data dataset (7 healthy, MCI and AD subjects).

{\bf Registration method.} The ANTS non-linear registration algorithm was used,
as it performed best in the ADNI1 cross-validation experiment (see Results).

{\bf Label fusion.} Majority vote was used performed equally well in the ADNI1
cross-validation experiment and is the easiest label fusion method to operate.

{\bf Evaluation.}
Since the hippocampal segmentation protocols differ between the ADNI labels and
Winterburn atlases, this poses a problem for direct similarity comparisons
between labels produced by MAGeT brain and the ADNI labels.  To evaluate the
performance of MAGeT brain, we compare the correlation of MAGeT segmentation
volumes with manual segmentation (SNT) volumes. Additionally, we correlate the
hippocampal volumes of established automated segmentation methods to MAGeT
segmentations.

Additionally, we compared classification accuracy of subjects by diagnosis based
on hippocampal volume using both the SNT labels and our produced labels.

@@

@@

@@

{\bf Dataset evaluated.} To validate that MAGeT performance generalises to
across disease categories, we measure the performance of MAGeT on the
Schizophrenic First Episode Patient (SFEP) dataset.

{\bf Atlas and template libary.} - two different atlas sets: a manual
hippocampal segmentation of patients, and Winterburn atlas set.  
\todo{complete}

{\bf Registration method.} ANTS.
\todo{complete}

{\bf Label fusion.} Majority vote.
\todo{complete}

{\bf Evaluation.} We validate the FEP-atlas segmentations using Dice's Kappa,
and the Winterburn-atlas segmentations by correlating volumes. 
\todo{complete}

%%%%%
%%%%%              Experiments: Winterburn Atlases Validation              %%%%%
%%%%%
\subsubsection{Winterburn Atlases Validation}
{\bf Dataset evaluated.} - T1 BRAVO scans of the same subjects included in the 
Winterburn atlas set.  These scans are taken within .... weeks of the scans for 
the Winterburn atlases.
\todo{complete}

{\bf Atlas and template libary.} - Atlas library is Winterburn T1 atlases.  
Template library consists of all five T1 BRAVOs, plus 15 T1 healthy control images.
\todo{complete}

{\bf Registration method.} ANTS
\todo{complete}

{\bf Label fusion.} Majority vote.
\todo{complete}

@@

Segmentation accuracy is judged by difference in hippocampal volume.
\todo{complete}
